Lossy compression of DNA sequencing quality data

### 2015-04-15 manuscript/q-compression.tex
Some refinement of the text and figure legends.

\begin{figure*}[!tb]%figure2
%\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
\centerline{\includegraphics[width=7in]{compression_results.pdf}}
\caption{Mean squared error versus bits/base-call for different
  compression methods applied to the \textit{Rhodobacter sphaeroides}
  2.4.1, and \textit{Homo sapiens} chromosome 14 fragment libraries,
  and \textit{Escherichia coli} str. K-12 MG1655, and \textit{Mus
    musculus} data sets. 2B --- 2-bin encoding; P$n$ --- profiling
  with $n$ profiles; R$n$ --- modeling with polynomial regression
  models of degree $n$; Q$n$ --- QualComp with rate parameter of
  $n$. Dashed line corresponds to lossless Bzip2 compression of
  quality values, and dotted line corresponds to original uncompressed
  data.}
\label{fig:mse_vs_bpbp}
\end{figure*}

@@

We compare the \textsc{mse} versus bits/base-call of the
\textit{Rhodobacter sphaeroides} 2.4.1, \textit{Homo sapiens}
chromosome 14, \textit{Escherichia coli} str. K-12 MG1655, and
\textit{Mus musculus} data sets (Figure \ref{fig:mse_vs_bpbp}). We
only include the fragment libraries for the \textit{Rhodobacter
  sphaeroides} 2.4.1, and \textit{Homo sapiens} chromosome 14 data
sets, but the additional short-jump library results are available in
the supplementary (TODO:TABLE). Storing the uncompressed quality
values requires 1 byte per base-call because they are stored in
\textsc{ascii} format and is denoted by the dotted black asterisk in the
figure.  The lossless compression of each data set using BZip2 ranges
from 2.19 - 3.10 bits/base-call and is denoted by the colored asterisks on the x-axis.  The compression methods tend to cluster together across the
different data sets.  Across all data sets, the 0-degree polynomial
regression, profile encodings, and QualComp have the lowest
bits/base-call.

@@

The majority of compression methods retain more base-pairs after preprocessing than the uncompressed sequences (Figure \ref{fig:preprocessing}).
In general, as a given compression model increases in complexity, i.e., as the number of profiles, polynomial degrees, or rate increases, the amount of base-pairs kept approaches the number base-pairs kept using the uncompressed sequences.
The compression methods on the \emph{Mus Musculus} data set have the greatest proportion of retained base-pairs compared to the uncompressed sequences.
The \textit{Escherichia coli} str. K-12 MG1655 MiSeq data set has the smallest range.

@@

\section*{Acknowledgement}
We thank the organizers and participants of the 2014 University of
T\"{u}bingen and University of Maryland Bioinformatics Exchange for
Students and Teachers (BEST) Summer School.

>>>>

\begin{figure*}[!tb]%figure2
%\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
\centerline{\includegraphics[width=7in]{compression_results.pdf}}
\caption{Mean squared error versus bits/base-call for different
  compression methods applied to the \textit{Rhodobacter sphaeroides}
  2.4.1, and \textit{Homo sapiens} chromosome 14 fragment libraries,
  and \textit{Escherichia coli} str. K-12 MG1655, and \textit{Mus
    musculus} data sets. 2B --- 2-bin encoding; P$n$ --- profiling
  with $n$ profiles; R$n$ --- modeling with polynomial regression
  models of degree $n$; Q$n$ --- QualComp with rate parameter of
  $n$. Asterisks denote the corresponding lossless compression using
  Bzip2, with the black asterisk corresponds to original uncompressed
  data.}
\label{fig:mse_vs_bpbp}
\end{figure*}

@@

We compare the \textsc{mse} versus bits/base-call of the
\textit{Rhodobacter sphaeroides} 2.4.1, \textit{Homo sapiens}
chromosome 14, \textit{Escherichia coli} str. K-12 MG1655, and
\textit{Mus musculus} data sets (Figure \ref{fig:mse_vs_bpbp}). We
only include the fragment libraries for the \textit{Rhodobacter
  sphaeroides} 2.4.1, and \textit{Homo sapiens} chromosome 14 data
sets, but the additional short-jump library results are available in
the supplementary (TODO:TABLE). Storing the uncompressed quality
values requires 1 byte per base-call because they are stored in
\textsc{ascii} format and is denoted by the dotted black asterisk in
the figure.  The lossless compression of each data set using BZip2
ranges from 2.19 - 3.10 bits/base-call and is denoted by the colored
asterisks on the abscissa. The compression methods tend to cluster
together across the different data sets. Across all data sets, the
0-degree polynomial regression, profile encodings, and QualComp have
the lowest bits/base-call.

@@

The majority of compression methods retain more base-pairs after
preprocessing than the uncompressed sequences (Figure
\ref{fig:preprocessing}).  In general, as a given compression model
increases in complexity, i.e., as the number of profiles, polynomial
degrees, or rate increases, the amount of base-pairs kept approaches
the number base-pairs kept using the uncompressed sequences.  The
compression methods on the \textit{Mus Musculus} data set have the
greatest proportion of retained base-pairs compared to the
uncompressed sequences.  The \textit{Escherichia coli} str. K-12
MG1655 MiSeq data set has the smallest range.

@@

\section*{Acknowledgement}
We thank the organizers and participants of the 2014 Bioinformatics
Exchange for Students and Teachers (BEST) Summer School, which was
funded by the offices of the Dean of The Graduate School at University
of Maryland and the Rektor of University of T\"{u}bingen.
